---
title: "RankedCoalescent_Example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RankedCoalescent_Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


# Overview 

The "RankedCoalescent" package provides tools for working with ranked, unlabeled,
rooted bifurcating tree shapes. It embeds the space of these tree shapes in a
Markov chain, and using these methods, one can compute Fréchet mean trees, 
perform statistical tests of neutral coalescence/branching, and simulate trees
from either the Blum-Francois or structured coalescent model.

# Load package

```{r setup}
library(RankedCoalescent)
library(fmatrix) #needed for plotting the trees. fmatrix::plotF.list() gives error
```

# Set up variables

```{r}
binM25 <- load_binM25()
diffM25 <- load_diffM25()
ndiff25 <- load_ndiff25()
```

# Generate state space and transition probabilities

```{r}
n <- 15
RankedCoalSeq <- ranked_coal_seq(n)
StSpM <- RankedCoalSeq$StSpM
SeqProbs <- RankedCoalSeq$seqProbs
```

# Computing moments

```{r}
Tiers <- make_Tiers(StSpM)
IndexList <- non_fixed_indices(n)
RewardMatrixF <- reward_matrix_F(StSpM, Tiers, IndexList)
InitDist <- initial_distribution(StSpM)
LeftProducts <- left_products_seq(InitDist, SeqProbs, RewardMatrixF, Tiers, IndexList)
RightProducts <- right_products_seq(SeqProbs, RewardMatrixF, Tiers, IndexList)
#Compute MeanTree and mean and covariance of non-fixed entries
MeanTree <- mean_tree_DPH(n, LeftProducts, IndexList)
MeanTreeNoFixed <- mean_tree_DPH_nofixed(LeftProducts, IndexList)
CovMat <- covariance_matrix_F(LeftProducts, RightProducts, IndexList, RewardMatrixF)
```


# Find Kingman Fréchet means for n = 15

```{r}
#Compute cost tiers and antecedents
CostTiers <- make_cost_tiers(Tiers, StSpM, MeanTree)
AntecedentList <- ranked_coal_list(n,TRUE)$AntecedentList
FréchetCostAndTrees <- ViTreebi(n, AntecedentList, StSpM, CostTiers, Tiers)
Cost <- FréchetCostAndTrees$MinCost
FréchetFmat1 <- path_to_Fmat(FréchetCostAndTrees$FrechetPaths[[1]], StSpM) 
FréchetFmat2 <- path_to_Fmat(FréchetCostAndTrees$FrechetPaths[[2]], StSpM)
#Visualize Fréchet means
plotF.list(list(FréchetFmat1,FréchetFmat2))
```

# Generate a sample of 1000 trees from the Blum-Francois model and Island model

```{r}
set.seed(42)
#Sample trees from Blum-Francois model and Island model
FsampleFromBF <- gen_Fmats_from_BF(-0.2, m = 1000, n)
FsampleFromIslandModel <- rRankedIslandCoal(m = 1000, n, K = 2, migrationRate = 0.5, startDist = c(12, 3))
```

# Test neutrality via non fixed entries

```{r}
eigenSigma <- eigen(CovMat)
Sigma0invsqrt <- eigenSigma$vectors %*% diag(sqrt(1/eigenSigma$values)) %*% t(eigenSigma$vectors) 
wald_test_F(FsampleFromBF, MeanTreeNoFixed, Sigma0invsqrt, IndexList)
wald_test_F(FsampleFromIslandModel, MeanTreeNoFixed, Sigma0invsqrt, IndexList)
```

# Fréchet mean(s) of sampled trees

```{r}
MeanBF <- meanF_from_sample(FsampleFromBF)
MeanIsland <- meanF_from_sample(FsampleFromIslandModel)
CostTiersBF <- make_cost_tiers(Tiers, StSpM, MeanBF)
CostTiersIsland <- make_cost_tiers(Tiers, StSpM, MeanIsland)

ViTreebiBF <- ViTreebi(n, AntecedentList, StSpM, CostTiersBF, Tiers)
ViTreebiIsland <- ViTreebi(n, AntecedentList, StSpM, CostTiersIsland, Tiers)

nFrechetBF <- length(ViTreebiBF$FrechetPaths)
nFrechetIsland <- length(ViTreebiIsland$FrechetPaths)
FrechetBF <- list() ; FrechetIsland <- list()
for(i in 1:nFrechetBF){FrechetBF[[i]] <- path_to_Fmat(ViTreebiBF$FrechetPaths[[i]], StSpM)}
for(i in 1:nFrechetIsland){FrechetIsland[[i]] <- path_to_Fmat(ViTreebiIsland$FrechetPaths[[i]], StSpM)}

plotF.list(FrechetBF)
plotF.list(FrechetIsland)
```


# Computing distribution of external branch length

```{r}
RankedBCP <- bcp_coal(n)
StSpMBCP <- RankedBCP$StSpM
ProbMBCP <- RankedBCP$ProbM
RewardE <- reward_E_BCP(StSpMBCP)
BCPDPH <- PhaseTypeR::DPH(ProbMBCP,c(1,rep(0,nrow(ProbMBCP)-1)))
EDPH <- DPH_E(BCPDPH, RewardE)
MinE <- min_E(n) ; MaxE <- max_E(n)
DistE <- dist_E(EDPH$subint_mat, EDPH$init_probs, MinE, MaxE)
plot(MinE:MaxE, DistE, type = "h")
points(MinE:MaxE, DistE, pch = 16, col = "red")
```

# Test neutrality based on external branch length

```{r}
binnedProbs <- binned_probs_E(length(FsampleFromBF), DistE)
Bins <- binnedProbs$Bins ; ExpectedBinned <- binnedProbs$ExpectedBinned
GoF_E(FsampleFromBF, Bins, ExpectedBinned, MinE, MaxE)
GoF_E(FsampleFromIslandModel, Bins, ExpectedBinned, MinE, MaxE)
```

# Test neutrality based on balance indices

```{r}
Mu0SE <- c(mean_S(MeanTreeNoFixed), mean_E(MeanTree))
Sigma0SE <- covariance_matrix_SE(CovMat, IndexList)
Sigma0SEInvSqrt <- eigen(Sigma0SE)$vectors %*% diag(1/sqrt(eigen(Sigma0SE)$values)) %*% t(eigen(Sigma0SE)$vectors)
wald_test_SE(FsampleFromBF, Mu0SE, Sigma0SEInvSqrt, IndexList)
wald_test_SE(FsampleFromIslandModel, Mu0SE, Sigma0SEInvSqrt, IndexList)
```

# Summary

The RankedCoalescent package provides tools for analysing ranked tree shapes.
This vignette demonstrated how to compute theoretical moments, simulate trees,
estimate Fréchet means, and perofrm neutrality tests based on several statistics.

